Refinements for open pNets

### 2021-07-26 2021-Refinement/paper/refinementarticle.tex
finally: V1 or deadlock preservation

\subsection{Open Automata}\label{sec:def}
 Open automata are labelled transition systems with variables  that can be used to compose other automata: they are made of transitions that are dependent of the actions of ``holes'', a composition operation consists in filling a hole with another automaton to obtain a more complete automaton. The variables makes the automata symbolic, and the holes allow for a partial definition of the behaviour.

\begin{definition}[Open transition, Open automaton]
An \emph{open automaton} is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the finite set of variable names, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.

@@

Open automata are partially specified automata, that partiality comes mostly from the holes.
A hole is an interface in which we can plug an open automaton.
The plugging operation is called composition.
The composition of open automata was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never completely formalised on open automata.
The definition of composition below is a direct translation of what happens with pNets composition without the need of introducing pNets.
\begin{definition}[Composition of open automata]
The composition of \(A_c = \OAg[c]\) in the hole \(k \in J_p\) of \(A_p = \OAg[p]\) is the open automaton defined as follows.
\[
	A_p\subst{A_c}{k} ::=  \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \]

\begin{align*}
	\text{with } T = & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{ \OT{s_p}{s'_p}{\alpha_p}{\beta_j^{j \in J'_p \uplus \mbrc{k} }}{g_p}{\psi_p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p} \in T_p, k \notin J'_p, s_c \in S_c}
\end{align*}
\end{definition}
The first automaton decides when the second can evolve by involving its hole in a transition:
The action emitted when \(A_c\) makes a transition is synchronised with the action of the hole \(k\) in transitions of \(A_p\) .

@@

Similarly to FH-bisimulation we are interested here in finding relations between states of to open automata that contain variables and holes. However here we want to biuld a refinement relation that is a simulation but also guarantees that no deadlock is introduced.
Our objective is to characterize when a relation between two states of two open automata have the desired property. Such a relation has the following signature: \( S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
We first define when a relations is deadlock reducing, before characterising a first simple refinement relation only valid to compare automata with identical holes.

\begin{definition}[Deadlock reduction]\label{def:dpwd}
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if  it satisfies the following:
\[ \forall \mpar{s_1, s_2} \in S_1 \times S_2, V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \]
\end{definition}

@@

Before characterising a composition which does not introduce deadlocks we need to be able to tell which states are reachable.
\begin{definition}[Reachability]
For any open automata \(A = \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is any predicate on states that is preserved accross transitions, and valdi on initial state:
\[\sigma_0 \vdash \reach{A}\mpar{s_0}\quad\land\quad\forall t = \OTg \in T, \fvars{t} \vdash \reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}\]
\end{definition}
Reachability takes into account all paths, and potentially over-approximate the reachable configurations and must be representable as a formula.
% LUDO HERE
\begin{definition}[Non-locking composition, intuitive definition]
Let \(A_i = \OAg[i], 0 \leq i \leq n\).
The composition \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is non-locking if \(A\) has a reachability predicate satisfying:
From reachable configurations, if there was a possible transition in \(A_0\) then there is a possible transition in \(A\).
Formally for \(\OAg = A\), it gives
\begin{multline*}
	\forall s = \mpar{s'_i}_{0 \leq i \leq n} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A}\mpar{s}} \implies \\
	\mpar{\exists t_0 \in \fOT{s'_0}, \exists \nu_0: \fvars{t_0} \to \values, \sigma \uplus \nu_0 \vdash \fguard{t_0}} \implies \\
	\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}
\end{multline*}
\end{definition}
As in the previous section, the definition can be simplified into a less intuitive version:
\begin{definition}[Non-locking composition, working definition]
\(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) is a non-locking composition if:
\[ \forall s = \mpar{s_0, s'} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{definition}
\begin{lem} The intuitive and working definition of non-locking composition are equivalent. \end{lem}
We want to prove
\[ \forall s = \mpar{s_p, s_c} \in S, V \uplus \biguplus_\subbox{t_p \in \fOT{s_p}} \fvars{t_p} \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{s_p}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \tag{WD}\label{eq:nlcWD} \]
\[ \bigsymb{\iff} \]
\begin{multline*}
	\forall s = \mpar{s_p, s_c} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A}\mpar{s}} \implies \\
	\mpar{\exists t_p \in \fOT{s_p}, \exists \nu: \fvars{t_p} \to \values, \sigma \uplus \nu \vdash \fguard{t_p}} \implies \\
	\mpar{\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}} \tag{ID}\label{eq:nlcID}
\end{multline*}

Hole - = relation

>>>>

\subsection{Open Automata}\label{sec:def}
 Open automata (abbreviated OA) are labelled transition systems with variables  that can be used to compose other automata: they are made of transitions that are dependent of the actions of ``holes'', a composition operation consists in filling a hole with another automaton to obtain a more complete automaton. The variables makes the OA symbolic, and the holes allow for a partial definition of the behaviour.

\begin{definition}[Open transition, Open automaton (OA)]
An \emph{open automaton} is a tuple \(\OAg\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the finite set of variable names, \(\sigma_0: V' \to \values\) the initial valuation of variables where \(V' \subseteq V\), \(J\) the set of hole names and \(T\) the set of open transitions.

@@

Open automata are partially specified automata, that partiality comes mostly from the holes.
A hole is an interface in which we can plug an open automaton.
The plugging operation is called composition.
The composition of open automata was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never completely formalised on open automata.
The definition of composition below is a direct translation of what happens with pNets composition without the need of introducing pNets.
\begin{definition}[Composition of open automata]
The composition of \(A_c = \OAg[c]\) in the hole \(k \in J_p\) of \(A_p = \OAg[p]\) is the OA defined as follows.
\[
	A_p\subst{A_c}{k} ::=  \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \]

\begin{align*}
	\text{with } T = & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{ \OT{s_p}{s'_p}{\alpha_p}{\beta_j^{j \in J'_p \uplus \mbrc{k} }}{g_p}{\psi_p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p} \in T_p, k \notin J'_p, s_c \in S_c}
\end{align*}
\end{definition}
The first OA decides when the second can evolve by involving its hole in a transition:
The action emitted when \(A_c\) makes a transition is synchronised with the action of the hole \(k\) in transitions of \(A_p\) .

@@

Similarly to FH-bisimulation~\cite{fhbisim} we are interested here in finding relations between states of to open automata that contain variables and holes. However here we want to build a refinement relation that is a simulation but also guarantees that no deadlock is introduced.
Our objective is to characterize when a relation between two states of two open automata have the desired property. Such a relation has the following signature: \( S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
We first define when a relations is deadlock reducing, before characterising a first simple refinement relation only valid to compare automata with identical holes.

\subsection{Deadlock Preservation, Reachability, and Composition}

\begin{definition}[Deadlock reduction]\label{def:dpwd}
Let \(\OAg[1]\), \(\OAg[2]\) be two OAs.
A relation on OA configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if  it satisfies the following:
\begin{multline*}
 \forall \mpar{s_1, s_2} \in S_1 \times S_2,\\ V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash \left(R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1} \right)
\end{multline*}
\end{definition}

\TODO{explain in particular why guards do the trick}

@@

Before characterising a composition which does not introduce deadlocks we need to be able to tell which states are reachable.
\begin{definition}[Reachability]
For any open automata \(A = \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is any predicate on states that is preserved across transitions, and valdi on initial state:
\[\sigma_0 \vdash \reach{A}\mpar{s_0}\quad\land\quad\forall t = \OTg \in T, \fvars{t} \vdash \left(\reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}\right)\]
\end{definition}
Reachability takes into account all paths, and potentially over-approximate the reachable configurations and must be representable as a formula.
% LUDO HERE
\begin{definition}[Non-locking composition]
Let \(A_i = \OAg[i], 0 \leq i \leq n\).
The composition \(A = A_0\mdbrk{j_i \mapsto A_i \middle| 1 \leq i \leq n}\) 
 is a non-locking composition if if \(A \) has a reachability predicate satisfying the following.
From reachable configurations, if there was a possible transition in \(A_0\) then there is a possible transition in \(A\):
\[ \forall s = \mpar{s_0, s'} \in S, V \uplus \biguplus_\subbox{t_0 \in \fOT{s_0}} \fvars{t_0} \vdash \left(\reach{A}\mpar{s} \wedge \bigvee_\subbox{t_0 \in \fOT{s_0}} \fguard{t_0} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \right)\]
\end{definition}
Again we use guards as a characteristics that the transition occurs.

\subsection{Refinement Relations for automata with the same holes}
\TODO{Hole - = relation}
